---
title: "Australia Bushfire Prediction"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE, echo=FALSE}
library(randomForest)
library(httr)
library(jsonlite)
library(lubridate)
library(dplyr)
```



Predicting the possibilities of a bushfire is one of the main aim of providing this dashboard. Given a probability of whether a bushfire is likely to occur near your place of residence, will provide sufficient time to make necessary arrangements to protect yourself, property and maybe wildlife. Here below, lies our predictive model, created and trained using a Random Forest machine learning algorithm, we are able provide the likelihood of bushfire occurrences. 

To use our software, please type in the postcode of your residence in the specified format. Only an Australian postcode is accepted and if an invalid postcode is given, an error will occur. Once a valid postcode is provided, the model will return the likelihood of bushfires with the suburb name and date.


```{r postcode_predictive_model, echo=FALSE}
textInput(inputId = "postcode", label = "Postcode (E.g. 3168)", value = NULL, width = NULL, placeholder = NULL)
actionButton("predictButton", "Predict", class = "btn-success")
set.seed(123456) 
#read bushfire data
bushfire.omit = read.csv("bushfire.csv")
#convert columns to the correct data type
bushfire.omit = bushfire.omit[,!(names(bushfire.omit) %in% c("Date"))]
bushfire.omit$Month = as.factor(bushfire.omit$Month)
for (i in 1:ncol(bushfire.omit)){
  if (class(bushfire.omit[,i]) == "character"){
    bushfire.omit[,i] = as.factor(bushfire.omit[,i])
  }
  if (class(bushfire.omit[,i]) == "integer"){
    bushfire.omit[,i] = as.numeric(bushfire.omit[,i])
  }
}
train.row = sample(1:nrow(bushfire.omit), 0.7*nrow(bushfire.omit))
bushfire.train = bushfire.omit[train.row,]
bushfire.test = bushfire.omit[-train.row,]
bushfire.fit.rf.ver2 = randomForest(Occurance ~ StateTerritory + MaxTemp + Month + WindDir9am + Temp3pm + Pressure9am + Humidity3pm ,data = bushfire.train, ntree = 90, mtry = 3)
#read direction table
directions = read.csv("directions.csv")
directions$cardinal = as.factor(directions$cardinal)
#----prediction----
#this function allow users to enter an Australia postcode and returns the
#forecast of weather metrics needed for the predictive model 
#param: Australia postcode
#return: data frame containing forecasted weather data and variables for the predictive model
weather.forecast = function(postcode){
  #if postcode is not valid, function return 0
  if((is.na(as.numeric(postcode))) | (nchar(postcode) != 4)){
    return(0)
  }
  #API call to get the State of the respective location
  locationQuery = paste("http://v0.postcodeapi.com.au/suburbs/",postcode,".json", sep = "")
  locRes = GET(locationQuery)
  loc.api.data = fromJSON(rawToChar(locRes$content))
  
  #if API returns values, postcode is valid
  #else postcode is not valid and function return 0
  if(length(loc.api.data) != 0){
    
    #create factor data type
    stateTerritory = as.factor(c("Australian Capital Territory", "New South Wales", "Northern Territory", 
                                 "Queensland", "South Australia", "Tasmania", "Victoria", "Western Australia"))
    month = as.factor(c(1,2,3,4,5,6,7,8,9,10,11,12))
    
    st = stateTerritory[stateTerritory == loc.api.data$state$name[1]]
    
    #query using the API
    query = paste("http://api.openweathermap.org/data/2.5/forecast?appid=c2c972a8600b40eedf993d1041b66083&units=metric&zip=" ,postcode,",au", sep = "")
    res = GET(query)
    api.data = fromJSON(rawToChar(res$content))
    
    #get next day's forecast
    #loop through the forecasts returned and identify the next day's index
    for (i in 1:40) {
      if (day(api.data$list[1,]$dt_txt) != day(api.data$list[i,]$dt_txt)){
        target = i
        break
      }
    }
    
    #create new data frame to store forecast observations
    obs = data.frame(StateTerritory = st )
    
    #collect forecasted data for 3 day
    
    for (counter in 1:3){
      #the API will provide forecast for every 3 hours, hence there should have 
      #8 forecasts everyday hence this variable identifies the index of the last 
      #forescast of the day
      target.end = target + 7
      
      #get 9am and 3pm forecast values
      forecast.9am = api.data$list[target+3,]
      forecast.3pm = api.data$list[target+5,]
      
      #calculate max temp of the day
      maxtemp = max(api.data$list[target:target.end,]$main$temp_max)
      
      #get the month of the date
      m = month[month == month(forecast.9am$dt_txt)]
      
      #get wind direction
      windDir = directions[forecast.9am$wind$deg >= directions$degree_min & forecast.9am$wind$deg < directions$degree_max,]$cardinal
      
      #get temp at 3pm
      temp3pm = forecast.3pm$main$temp
      
      #get pressure at 9am
      pressure9am = as.numeric(forecast.9am$main$pressure)
      
      #get humidity at 3pm
      humidity3pm = as.numeric(forecast.3pm$main$humidity)
      
      #compile all values into a data frame
      
      #if this is the variable values for the first day
      #create new columns for values
      #if not, add the new values to the existing dataframe
      if (length(obs) == 1){
        obs$MaxTemp = maxtemp
        obs$Month = m
        obs$WindDir9am = windDir
        obs$Temp3pm = temp3pm
        obs$Pressure9am = pressure9am
        obs$Humidity3pm = humidity3pm
        obs$Location = api.data$city$name
        obs$Date = as.Date(forecast.9am$dt_txt)
      }
      else{
        new.obs = data.frame(StateTerritory = st )
        new.obs$MaxTemp = maxtemp
        new.obs$Month = m
        new.obs$WindDir9am = windDir
        new.obs$Temp3pm = temp3pm
        new.obs$Pressure9am = pressure9am
        new.obs$Humidity3pm = humidity3pm
        new.obs$Location = api.data$city$name
        new.obs$Date = as.Date(forecast.9am$dt_txt)
        
        obs = rbind(obs,new.obs )
      }
        
      
      
      #set target index to the next day
      target = target.end + 1
    }
    
    
    
    #return the data frame
    return(obs)
  }
  else{
    #return 0 if input is not valid
    return(0)
  }
  
  
}
#use forecast data from API to predict fire occurance in Austarlia
#param: Australian postcode
#return: data frame with prediction result
fire.prediction = function(postcode){
  
  #use forecast.value function to make API call
  forecast.value = weather.forecast(postcode)
  
  #make prediction using predictive model
  if(is.data.frame(forecast.value)){
    prediction = predict(bushfire.fit.rf.ver2, forecast.value[1:7])
    #create and format result table for output
    result = cbind(forecast.value[1], forecast.value[8:9], Prediction = prediction)
    colnames(result) = c("State/Territory", "Location", "Date","Prediction")
    result$Prediction = as.character(result$Prediction)
    
    if (length(result[result$Prediction == "No",]$Prediction) != 0){
      result[result$Prediction == "No",]$Prediction = "Low Chance"
    }
    if (length(result[result$Prediction == "Yes",]$Prediction) != 0){
      result[result$Prediction == "Yes",]$Prediction = "High Chance"
    }
    
    
  }
  
  else{
    result = "Invalid input (Please try again with a valid postcode)"
    
  }
  
  #return result table
  result
  return(result)
  
}

# set the function of the button
result <- eventReactive(input$predictButton, fire.prediction(input$postcode))

# print the result
renderPrint({
  result()
})
  
```



Click the button below to access the visualize dashboard:

```{r back-button, echo = FALSE}
# A button to access the visualize dashboard

actionButton(inputId = "backButton", label = "Back To Main Dashboard", icon = icon("angle-double-left"), onclick ="window.open('https://fit3164-team15-s2-2020.shinyapps.io/Dashboard')")


```






